---
title: "Hazard Models"
author:
- Ozlem Tuncel^[Georgia State University, otuncelgurlek1@gsu.edu], updated: Patrick
    Munger
date: "Fall 2024"
output:
  html_document:
    df_print: paged
urlcolor: blue
editor_options:
  markdown:
    wrap: 72
---


*Survival/Hazard/Duration/Event History models are complex and are courses of there own in some
disciplines and programs. We'll review the basics, but here are some
more in depth resources:*

[Event History Modeling: A Guide for Social Scientists by
Box-Steffensmeier and
Jones](https://www.amazon.com/Event-History-Modeling-Scientists-Analytical/dp/0521546737?asin=0521546737&revisionId=&format=4&depth=1)

[Emily Zabor's
guideline](https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html)

[Great source of diagnostic
interpretation](https://shariq-mohammed.github.io/files/cbsa2019/1-intro-to-survival.html#61_model_fit)
[Interpreation](https://stats.oarc.ucla.edu/stata/seminars/stata-survival/)
[Interpreation2](https://bioconnector.github.io/workshops/r-survival.html#kaplan-meier_plots)

[Hazard Ratio
interpretation](http://onbiostatistics.blogspot.com/2020/05/interpreting-hazard-ratio-can-we-say.html)
[Parametric
Nonparametric](https://www.mayo.edu/research/documents/parametric-and-nonparametric-demystifying-the-terms/doc-20408960)
[Survival
basics](http://www.sthda.com/english/wiki/survival-analysis-basics)
[Parametric
nonparametric](https://boostedml.com/2018/11/when-should-you-use-non-parametric-parametric-and-semi-parametric-survival-analysis.html)

# Agenda
1.  Load [breast cancer
    data](https://rdrr.io/cran/survival/man/rotterdam.html) from
    Rotterdam tumor bank
2.  R script to run Cox Proportional Hazards Model
3.  Test assumptions
4.  Weibull and Exponential
5.  Replication trial

# Preliminaries

```{r set_seed, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Specify any integer
set.seed(1234)
```

```{r}
library(tidyverse) ## Utility tools
library(ggsurvfit) # Plotting
library(survival) # Base survival models
library(survminer) # Some graphics support
library(stargazer) ## Tables
library(car) # Companion to Applied Regression
library(carData) # Supplemental Data
library(gmodels) # Crosstabs
library(Formula)
library(reshape2)
library(eha) # Extensions of survival
```

# Upload data

We are using a dataset on [breast
cancer](https://rdrr.io/cran/survival/man/rotterdam.html).

For `survival` package, we need to first identify key variables for our
analysis, specifically the time variable and the event or
failure variable. 

In this case, our time variable is `dtime`, which is a count of days from the start of the trial to death or last follow up. Our survival variable is `death`, a binary indicating if the patient dies during the trial. 

We use *Surv()* function to identify these variables and create an outcome, we'll call `status`. 

```{r}
rott <- survival::rotterdam # Loading in the data
status <- Surv(rott$dtime, rott$death) # Creating an 'outcome' variable

str(rott)
rott$size <- as.numeric(rott$size) # Data cleaning 
```

# Cox PH Models

```{r}
cox_m1 <- coxph(status ~ chemo + size + age, data=rott)
summary(cox_m1)

```


Variables with positive coefficients are associated with increased
hazard and decreased survival times, i.e. as the predictor increases the
hazard of the event increases and the predicted survival duration
decreases. Negative coefficients indicate decreased hazard and increased
survival times. In our case chemo treatment, size of the tumor, and age
increase the hazard and decrease the survival time.

Why would chemo decrease survival times in our sample?

HR=1 : No effect HR\>1: Increase in hazard HR\<1: Decrease in hazard

The exp(coef) column contains the hazard ratio – the multiplicative
effect of that variable on the hazard rate. For instance, one unit
increase in the tumor size leads to 1.8-fold increase in hazard.

# Graph the whole model

A Kaplan-Meier curve is an estimate of survival probability at each
point in time (given survival up until that point). It has very few
assumptions and is a purely descriptive method. However, it does not
provide any information on the effect of covaraites.

We can produce it with the `survfit()` function, adding confidance bands and some other information in the second plot. 

```{r}
survfit(cox_m1, data = rott) %>% 
  ggsurvfit() 

# Let's make it fancier
survfit(cox_m1, data = rott) %>% 
  ggsurvfit() +
  labs(x = "Days to death or last follow-up",
       y = "Overall survival probability") +
  scale_x_continuous(breaks = seq(0, 7043, by = 365)) +
  add_confidence_interval() +
  add_risktable()
```
At risk: how many patients are still alive and un-censored at that point.
Events: cumulative deaths at that point. 

# Using survplot to see how covariates affect the outcome

Creating data for survplot

```{r}
chemo_data <- with(rott,                            
               data.frame(age = rep(mean(age),2),
                          chemo = c(0, 1),
                          size =  c(2, 2)))
```

Plotting to asses the effects of chemo

```{r}
ggsurvplot(survfit(cox_m1, newdata = chemo_data), 
           data = chemo_data,
           palette = "uchicago",
           surv.median.line = "hv",
           conf.int = T,
           legend.labs = c("Chemo not received", "Chemo received"),
           break.x.by = 365,
           break.y.by = 0.1)
```

If we take chemo treatment into account we see that after a certain
point in time, there is a clear divergence in terms of survival
probability.


# Let's take age and chemo into account by ploting survival probabilities for 25 and 75 year olds, with and without chemo, holding tumor size at its median. 

```{r}
age_data <- with(rott,
                 data.frame(age = c(25, 25, 75, 75),
                             chemo = c(0, 1, 0, 1),
                             size = rep(mean(size), 4)))

ggsurvplot(survfit(cox_m1, newdata = age_data), 
           data = age_data, 
           palette = "uchicago",
           conf.int = T,      
           legend.labs = c("Age=25\nChemo=0", "Age=25\nChemo=1", 
                           "Age=75\nChemo=0", "Age=75\nChemo=1" ),
           break.x.by = 365,
           break.y.by = 0.1,
           surv.median.line = "hv")
```

Age matters a lot in probability of survival from breast cancer. Older
people have lower chance of living with or without chemo treatment
compared to younger people.

Let's take tumor size into account, holding age at its mean.

```{r}
size_data <- with(rott,                            
                  data.frame(age = rep(mean(age),6),
                             chemo = c(0, 0, 0, 1, 1, 1),
                             size =  c(1, 2, 3, 1, 2, 3)))

# Creating the survplot
ggsurvplot(survfit(cox_m1, 
                   newdata = size_data), 
           data = size_data, 
           conf.int = T,  
           palette = "uchicago",
           legend.labs = c("Chemo=0\nSize=1", "Chemo=0\nSize=2", "Chemo=0\nSize=3",
                         "Chemo=1\nSize=1", "Chemo=1\nSize=2", "Chemo=1\nSize=3"),
           break.x.by = 365,
           break.y.by = 0.1,
           surv.median.line = "hv")
```

Size of the tumor matter a lot as well. Bigger tumors are likely to lead
to lower survival probability.

# Checking the proportional hazard assumption

In principle, the Schoenfeld residuals are independent of time. Coviariats are assumed to proportionally affect the hazard rate over time. A plot
that shows a non-random pattern against time is evidence of violation of
the PH assumption. If a covariate breaks the assumption, it is time dependent. 

We can test this post-estimation with `cox.zph()`. The null is that the assumption is met. 
So, a rejection of the null (p \< 0.05), suggests that the proportional hazards assumption
is violated for that covariate. 

```{r}
cox.zph(cox_m1)
```

Systematic departures from a horizontal line are indicative of
non-proportional hazards, since proportional hazards assumes that
estimates β1,β2,β3 do not vary much over time. We can visualize this with Schoenfeld plots:

```{r}
ggcoxzph(cox.zph(cox_m1))
```

# Picking a distribution


"Weibull distribution might appear curvilinear in the log versus time
plot, but will be linear in a loglog plot ln[− ln S(t)]"

"Exponential distribution will appear linear in both plots, and have a
slope equal to 1 in the loglog plot"

## Log(survival time) against time:

```{r}
sfit1 <- survfit(cox_m1) 
sum1 <- summary(sfit1, times = rott$dtime)
f1 <- log(surv) ~ (time)
t1 <- as.data.frame(sum1[c("time", "surv")])

fit.lm1 <- lm(f1, t1)
plot(f1, t1)
abline(fit.lm1)
```

Derivation from linearity suggests that hazard rate may not be constant
over time (as assumed by exponential model)

## Log(-log(survival time) against log(time):

```{r}

sfit2 <- survfit(cox_m1) 
sum2 <- summary(sfit2, times = rott$dtime)
f2 <- log(-log(surv)) ~ log(time)
t2 <- as.data.frame(sum2[c("time", "surv")])

# Add a small offset to prevent extreme survival values from causing issues
t2$surv <- ifelse(t2$surv == 1, 0.999999, t2$surv)
t2$surv <- ifelse(t2$surv == 0, 0.000001, t2$surv)

fit.lm1 <- lm(f2, t2)
plot(f2, t2)
abline(-10, 1)
```


# Shifting to fully parameterized

Fully parameterized models specifiy the form of the baseline hazard. We can use the `survreg()` function and specify a distribution. 

```{r, warning=FALSE}
w1 <- survreg(status ~ chemo + size + age, data = rott, dist = "weibull")
e1 <- survreg(status ~ chemo + size + age, data = rott, dist = "exponential")

stargazer(w1, e1, type = "text")

stargazer(w1, e1 , 
          type = "text", 
          column.labels = c(w1$dist, e1$dist),
          add.lines=list(c("Log(scale)", signif(log(w1$scale),3), signif(log(e1$scale),3)),
                         c("Scale", signif(w1$scale,3), signif(e1$scale,3)),
                         c("")
          )
)
```

The shape parameter is not returned in the summary, but we can get it by
taking the inverse of the scale parameter:

```{r}
scale_param <- w1$scale
shape_param <- 1 / scale_param
shape_param
```

A shape parameter over 1 suggests an increasing hazard rate over time, under 1 suggests decreasing. 

## Comparison graph (from eha package)

Close alignment suggest Weibull is a good fit.

```{r}
status <- Surv(rott$dtime, rott$death) 
rott$size <- as.numeric(rott$size) 

# Cox regression 
cox <- coxreg(status ~ meno + grade + age, data = rott) 
# PH regression, standard is weibull
phw <- phreg(status ~ meno + grade + age, data = rott)  

check.dist(cox, phw) 
```

## Weibull version

The function rweibull(n, shape, scale) generates a sample of 10,000
observations from a Weibull distribution. When the shape parameter is
set to 1, it will converge to an exponential distribution.

```{r}
sample1 <- as.data.frame(rweibull(10000, 1, 1))
names(sample1) <- "Weibull"
head(sample1)

sample1 %>% 
  ggplot(aes(x = Weibull)) +
  geom_density(size = 1.2, color = "red") + 
  theme_minimal() + 
  labs(x = "Values", y = "Density") 

```

## Exponential version

```{r}
sample2 <- as.data.frame(rexp(10000, 1))
names(sample2) <- "Exponential"
head(sample2)

sample2 %>% 
ggplot(aes(x = Exponential)) +
  geom_density(size = 1.2, color = "blue") + 
  theme_minimal() + 
  labs(x = "Values", y = "Density") 
```







