---
title: 'Week 7: Dichotomous Predictors'
author: "Ozlem Tuncel; Patrick Munger"
date: "Spring 2025"
output: html_document
---

# Week 7: Dichotomous Predictors

## Agenda

-   Import and clean data
-   Estimate model with dichotomous IV
-   Compare plots of continuous and dichotomous predictors
-   Estimate model factoring out $k > 2$ categorical variable
    -   create our own dummies
    -   factor out in the model
-   Basic transformations

## Setup

### Load libraries:

There might be some new packages here. Make sure you have them all installed before loading them.

```{r}
library(tidyverse)  
library(stargazer)  
library(ggpubr)      
library(broom)      
library(PerformanceAnalytics) 
```

### Import and clean data

Here were are using the V-dem version 12 dataset again.

Remember to make sure your working directory is set to the directory where your data file is. Check current working with `getwd()` and change if necessary with `setwd("path to data directory goes here)`, replacing with the actual with to your data directory.

Note: when importing the data, only use /data if the file is in a directory within your working directory called "data". If your data file is located directly in your working ridectory simply specify the file name - "vdem12.rds"

```{r, warning=FALSE}
my_data <- readRDS("data/vdem12.rds") 
```

Now let's subset the data to US observation, select and rename some variables. For this week, we are going to introduce the polarization and regime variables. The regime variable has four possible categories but only three are present in the US data:

-   $1$: Electoral autocracy
-   $2$: Electoral democracy
-   $3$: Liberal democracy

The first polarization variable is an ordinal variable with four possible levels of polarization (there is a possible $0$ value for no polarization, but it is not present in the US data). The other polarization variable is a continuous interval with possible values between $-5$ and $5$ that is computed from the ordinal variable using a model that accounts for the reliability and response scale of the experts surveyed to generate the variable. More about this can be found on the V-dem methodology page linked [here](https://www.v-dem.net/about/v-dem-project/methodology/).

```{r}
us_data <- my_data |>
  filter(country_name == "United States of America") |> 
  rename(democracy = v2x_polyarchy, 
         gdp_per_capita = e_gdppc,
         urbanization = e_miurbani,
         regime = v2x_regime,
         polarization_ordinal = v2cacamps_ord,
         polarization = v2cacamps) |>
  dplyr::select(democracy, gdp_per_capita, urbanization, regime, polarization, polarization_ordinal)
```

For demo purposes, let's create a new variable that collapses the interval polarization variable into a binary of high polarization (values $\geq -1$ are converted to value $1$) and low polarization (values $\leq -1$ are converted to value $0$).

```{r}
us_data <- us_data |>
  mutate(high_polarization = ifelse(polarization >= -1, 1, 0))
```

### Plots

Before we estimate models, let's generate the plots from the slides that compare fitting an OLS line for a continuous and dichotomous variable. We will first plot GDP and democracy then our binary polarization variable and democracy. We can assign the plots to objects then put them together with the `ggarrange()` function.

```{r}
# Continuous IV 
cont <- us_data |> 
  ggplot(aes(x = gdp_per_capita, y = democracy)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  scale_x_continuous(breaks = seq(0, 80, 10)) +
  scale_y_continuous(breaks = seq(0, 2, 0.10)) +
  labs(x = "GDP per capita ($)", y = "Democracy") +
  theme_bw()

# Binary IV
binary <- us_data |> 
  filter(!is.na(high_polarization)) |> 
  ggplot(aes(x = high_polarization, y = democracy)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  scale_x_continuous(breaks = seq(0, 1, 1), limits = c(0, 1)) +
  scale_y_continuous(breaks = seq(0, 1, 0.10)) +
  labs(x = "High Political Polarization", y = "Democracy") +
  theme_bw()

# Combine 
ggarrange(cont, binary)
```

### Estimating models with dichotomous predictors.

Now let's estimate some models. First we will estimate a multiple OLS with our continuous GDP variable and our binary polarization variable as IVs. Since this polarization variable is already binary, it will produce a single coefficient estimate. Since high polarization is coded as $1$, we can interpret this coefficient as the effect of a country (or, in our case an annual observation of the US) being highly polarized (as compared to low polarization) on democracy.

```{r}
binary_model <- lm(democracy ~ gdp_per_capita + high_polarization, us_data)
summary(binary_model)
```

Now let's turn to our regime variable which has 3 categories ($k = 3$). These categories could perhaps be interpreted to have an order, but let's assume they are nominal. To estimate its effects on democracy then, we would want to factor it out into two separate dichotomous variables with one category as the reference ($0$). There are a couple ways we can do this. Firstly, we can manually split the variable. For the outcomes to be mutally exclusive and exhaustive, we will have to establish one category as the reference. Let's use category $1$ (electoral autocracy) as the reference and create dummies for categories $2$ (electoral democracy) and $3$ (liberal democracy).

We will call the first dummy `electoral democracy`, setting cateogry $2$ to value $1$, otherwise setting values to $0$. We will call the other dummy `liberal democracy`, setting category $3$ to value $1$, otherwise setting values to $0$.

We can use the `ifelse()` function and pipe the dummies into new columns in the df named according to the category that takes value of $1$.

```{r}
us_data$`electoral democracy` <- ifelse(us_data$regime == 2, 1, 0)  
us_data$`liberal democracy` <- ifelse(us_data$regime == 3, 1, 0)  
```

The other approach is more straightforward and can be done within the `lm()` function itself by simply wrapping our variable in `as.factor()`. If our categories are numerically coded, the lowest category will be set as the reference and the variable will be factored accordingly in the model, resulting in coefficient estimates for two separate dummies. Since we set category $1$ as our reference in the first approach, the results should be identical. Let's check.

In the first model, we separately specify the two dummies we manually created. In the second model, we specify only the regime variable with `as.factor()`.

-   note: theoretically it would also be silly to estimate the effects of regime on democracy as these are measuring very similar concepts, so they are obviously correlated, but the variable works well for demonstration purposes.

```{r, warning = FALSE}
mod1 <- lm(democracy ~ `electoral democracy` + `liberal democracy`, us_data)
mod2 <- lm(democracy ~ as.factor(regime), us_data)

stargazer(mod1, mod2, type = "text")
```

The results are the same! we can interpret the coefficients as the effects of a country (or an annual observation of the US) being an electoral democracy (as opposed to the other categories) on democracy and the effects of a country being a liberal democracy (as opposed to the other categories) on democracy. Again, this is a silly relationship theoretically since the concepts are so similar, but the intuition of the setup and interpretation is what is important here.

### Transformations

Let's look at the distributions and relationships between three of our variables with `chart.Correlation()` from the `PerformanceAnalytics` package.

```{r}
chart.Correlation(us_data |> dplyr::select(democracy, gdp_per_capita, urbanization))
```

To demonstrate how transformations of our variables can help the linearity of a relationship, let's focus on the relationship between gdp per capita and urbanization.

```{r, warning = FALSE}
us_data |> 
  ggplot(aes(x = gdp_per_capita, y = urbanization)) +
  geom_point() + 
  geom_smooth(method = "lm", color = "red") +
  geom_smooth() +
  scale_x_continuous(breaks = seq(0, 80, 10)) +
  scale_y_continuous(breaks = seq(0, 2, 0.10)) +
  labs(x = "GDP per capita", y = "Urbanization") +
  theme_bw()
```

Clearly there is a relationship, but it is not linear. Maybe more logarithmic. Let's try transforming GDP per capita by taking it's natural log then plotting. We can simply use the `log()` function for this.

```{r, warning = FALSE}
us_data |> 
  ggplot(aes(x = log(gdp_per_capita), y = urbanization)) +
  geom_point() + 
  geom_smooth(method = "lm", color = "red") +
  geom_smooth() +
  scale_x_continuous(breaks = seq(0, 80, 10)) +
  scale_y_continuous(breaks = seq(0, 2, 0.10)) +
  labs(x = "Logged GDP per capita", y = "Urbanization") +
  theme_bw()
```

It does not perfectly linearize the relationship but does improve it.

Let's compare the effects in a coefficient table.

```{r, warning = FALSE}
original <- lm(urbanization ~ gdp_per_capita, us_data)
logged <- lm(urbanization ~ log(gdp_per_capita), us_data)

stargazer(original, logged, type = "text")
```
