---
title: "Bivariate OLS"
author: "Ozlem Tuncel; Patrick Munger"
date: "Spring 2025"
output: html_document
---

# Week 4: Bivariate OLS

Implementation in R is relatively simple and straightforward this week.

### Agenda:

-   Import and clean data

-   Introduce function for estimating OLS: `lm()`

-   View model summary

-   Use stargazer to create well-formated tables

## Setup

### Load libraries:

We are introducing the `stargazer` package here. Make sure you have installed it with `install.packages("stargazer")` or in RStudio's packages pane before loading it.

```{r}
library(tidyverse) 
library(stargazer) # For well-formatted output tables (including latex code)
library(broom)
```

### Import and clean data

Here were are using the V-dem version 12 dataset again, the same we used for EDA in Week 2.

Remember to make sure your working directory is set to the directory where your data file is. Check current working with `getwd()` and change if necessary with `setwd("path to data directory goes here)`, replacing with the actual wath to your data directory.

```{r}
my_data <- readRDS("data/vdem12.rds")
```

For ease of interpretation, we will change the names of the two variables we will be using:

```{r}
my_data <- my_data |>
  rename(democracy = v2x_polyarchy, gdp_per_capita = e_gdppc)
```

# Estimate Bivariate OLS

To estimate an OLS model in R, we use the `lm()` (lm stands for linear model). Let's view the documentation for `lm()` using `?` followed by the function. We can also use `help(lm)`. This will list the various arguments that can be included in the function to modify how the model is estimated.

```{r}
?lm
```

For now, we will keep things simple and just specify the `formula` and `data`. The syntax is as follows, subtituting the actual names of your variables and data:

-   `lm(formula = dependent_variable ~ independent_variable, data = your_data)`

The dependent variable should always come first, followed by a tilda (`~`), then your independent variable. Place a comma after your independent variable, then specify the data from which the variables are pulled from. It is not necessary to include `data =`, as you can simply place your data name after the comma, but for demonstration purposes it is useful.

Note: it is not necessary to actually include `formula =` and `data =`, as R knows that the formula argument comes first and the data argument second, separated by a comma. Thus we could simply write:

-   `lm(dependent_variable ~ independent_variable, your_data)`

Now let's do this with our actual data and variables and estimate the model. We will use `democracy` as our dependent variable and `gdp_per_capita` as our independent variable.

```{r}
lm(democracy ~ gdp_per_capita, my_data) 
```

## Model Summaries

The output of running the model itself does not tell us that much. For more detailed summaries, we can assign the model to an object and then use `summary()`.

```{r}
my_lm <- lm(democracy ~ gdp_per_capita, my_data) 

summary(my_lm) 
```

Here we see that we now provided with summary stats of the residuals, uncertainty estimates alongside the coefficient values, and model statistics such as degrees of freedom, R-squared, etc.

Note: `<2.2e-16` is the precision limit R places on summaries of `lm()` models. When we see this, it means that the p-value is very, very close to zero.

## Format Regression Table

Now let's use the `stargazer` package to create a nicely formatted table that summarizes our model results. We use the function `stargazer()` (the same as the package name). First we will have it create a nice text-based table in the console.

```{r, warning = FALSE}
stargazer(my_lm, type = "text")
```

Now let's have it produce code that we can copy and paste into latex to produce a nice table. We simply change `type = "text"` to `type = "latex"`.

```{r, warning = FALSE}
stargazer(my_lm, type = "text")
```

We can copy and paste this output straight into latex to produce a article quality table.

We can add a bunch of other arguments to `stargazer()` to modify the table. Use `?stargazer` to view all possible arguments.

```{r, warning=FALSE}
stargazer(my_lm, 
          type = "text", # change to "latex" for latex code 
          title = "The relationship between democracy and GDP per capita", 
          covariate.labels = c("GDP per capita"),
          dep.var.labels = c("Electoral Democracy Index"),
          ci.level = 0.95,
          star.cutoffs = c(0.05),
          notes.align = "l",
          notes.append = FALSE,
          notes.label = "Notes", 
          notes = "*p < 0.05. Standard errors are in parentheses.")
```

Now let's make some modifications to remove some of the things that make Dr. Fix angry!

To make sure there are no abbreviations, it is easier to just make the modifications in your latex code, so change "Residual Std. Error" to "Residual Standard Error".

Additionally, even though we have explicitly told stargazer to round to 3 decimal places, numbers very close to 0 will still display as 0.0001 instead of 0.000, so we can change this in the latex code.

```{r, warning=FALSE}
stargazer(my_lm, 
          type = "text", # change to "latex" for latex code 
          title = "OLS Model: Democracy", 
          covariate.labels = c("GDP per capita"),
          dep.var.caption = "", # Remove the "Dependent Variable" caption
          dep.var.labels = "", # Remove the dependent variable label
          column.labels = "Coefficients", # Add "Coefficients" as the column label
          ci.level = 0.95,
          star.cutoffs = c(0.05),
          notes.align = "l",
          notes.append = FALSE,
          notes.label = "Notes", 
          notes = "*p < 0.05. Standard errors are in parentheses.",
          digits = 3) # Round coefficients to 3 decimal places

```

## Coefficient Plot

We can use the `tidy()` function from the `broom` package to extract the coefficient estimates and confidence intervals from the model then plot them with ggplot.

```{r}
model_tidy <- tidy(my_lm, conf.int = TRUE)
print(model_tidy)
```

Now plot with ggplot using `geom_errorbarh()` to include confidence intervals.

```{r}
ggplot(model_tidy, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  labs(title = "Coefficient Plot",
       x = "Coefficient Estimate") +
  theme_minimal()
```
