---
title: "Hazard Models"
author:
- Ozlem Tuncel^[Georgia State University, otuncelgurlek1@gsu.edu], updated: Patrick
    Munger
date: "Fall 2024"
output:
  html_document:
    df_print: paged
urlcolor: blue
editor_options:
  markdown:
    wrap: 72
---

# Today's agenda

*Cautionary notes* Usually this subject is a course of its own. So, we
will not be really able to cover everything in here. If you want to
focus on this sort of modelling, I highly recommend reading a book on
this. My recommendation:

[Event History Modeling: A Guide for Social Scientists by
Box-Steffensmeier and
Jones](https://www.amazon.com/Event-History-Modeling-Scientists-Analytical/dp/0521546737?asin=0521546737&revisionId=&format=4&depth=1)

Also, we will not be able to cover everything today. So, in case you
need further explanation, I recommend Emily Zabor's guide:

[Emily Zabor's
guideline](https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html)

[Great source of diagnostic
interpretation](https://shariq-mohammed.github.io/files/cbsa2019/1-intro-to-survival.html#61_model_fit)
[Interpreation](https://stats.oarc.ucla.edu/stata/seminars/stata-survival/)
[Interpreation2](https://bioconnector.github.io/workshops/r-survival.html#kaplan-meier_plots)

[Hazard Ratio
interpretation](http://onbiostatistics.blogspot.com/2020/05/interpreting-hazard-ratio-can-we-say.html)
[Parametric
Nonparametric](https://www.mayo.edu/research/documents/parametric-and-nonparametric-demystifying-the-terms/doc-20408960)
[Survival
basics](http://www.sthda.com/english/wiki/survival-analysis-basics)
[Parametric
nonparametric](https://boostedml.com/2018/11/when-should-you-use-non-parametric-parametric-and-semi-parametric-survival-analysis.html)

1.  [Today's breast cancer
    data](https://rdrr.io/cran/survival/man/rotterdam.html)
2.  R script to run Cox Proportional Hazards Model
3.  Test assumptions
4.  Replication trial

# Preliminaries

```{r set_seed, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Specify any integer
set.seed(1234)
```

```{r}
library(tidyverse) ## Utility tools
library(ggsurvfit) # Plotting
library(survival) # Base survival models
library(survminer) # Some graphics support
library(stargazer) ## Tables
library(car) # Companion to Applied Regression
library(carData) # Supplemental Data
library(gmodels) # Crosstabs
library(Formula)
library(reshape2)
library(eha) # Extensions of survival
```

# Upload data

We are using a dataset on [breast
cancer](https://rdrr.io/cran/survival/man/rotterdam.html).

For `survival` package, we need to first identify key variables for our
analysis. In this case this is the time variable and the event or
failure variable. We use *Surv()* function for this (STATA equivalent of
this command is stset).

```{r}
rott <- survival::rotterdam # Loading in the data
status <- Surv(rott$dtime, rott$death) # Creating an 'outcome' variable

summary(rott)
rott$size <- as.numeric(rott$size) # Data cleaning 
```

# Cox PH Models

```{r}
cox_m1 <- coxph(status ~ chemo + size + age, data=rott)
summary(cox_m1)

```

Variables with positive coefficients are associated with increased
hazard and decreased survival times, i.e. as the predictor increases the
hazard of the event increases and the predicted survival duration
decreases. Negative coefficients indicate decreased hazard and increased
survival times. In our case chemo treatment, size of the tumor, and age
increase the hazard and decrease the survival time.

HR=1 : No effect HR\>1: Increase in hazard HR\<1: Decrease in hazard

The exp(coef) column contains the hazard ratio – the multiplicative
effect of that variable on the hazard rate. For instance, one unit
increase in the tumor size leads to 1.8-fold increase in hazard.

# Graph the whole model

A Kaplan-Meier curve is an estimate of survival probability at each
point in time (given survival up until that point). It has very few
assumptions and is a purely descriptive method. However, it does not
provide any information on the effect of covaraites.

```{r}
survfit(cox_m1, data = rott) %>% 
  ggsurvfit() 

# Let's make it fancier
survfit(cox_m1, data = rott) %>% 
  ggsurvfit() +
  labs(x = "Days to death or last follow-up",
       y = "Overall survival probability") +
  scale_x_continuous(breaks = seq(0, 7043, by = 365)) +
  add_confidence_interval() +
  add_risktable()
```

# Using survplot to see how covariates affect the outcome

Creating data for survplot

```{r}
chemo_data <- with(rott,                            
               data.frame(age = rep(mean(age),2),
                          chemo = c(0, 1),
                          size =  c(2, 2)))
```

Plotting to asses the effects of chemo

```{r}
ggsurvplot(survfit(cox_m1, newdata = chemo_data), 
           data = chemo_data,
           palette = "uchicago",
           surv.median.line = "hv",
           conf.int = T,
           legend.labs = c("Chemo not received", "Chemo received"),
           break.x.by = 365,
           break.y.by = 0.1)
```

If we take chemo treatment into account we see that after a certain
point in time, there is a clear divergence in terms of survival
probability.

# Let's take age into account and create data for survplot and Creating the survplot

```{r}
age_data <- with(rott,
                 data.frame(age = c(25, 25, 75, 75),
                             chemo = c(0, 1, 0, 1),
                             size = rep(median(size), 4)))

ggsurvplot(survfit(cox_m1, newdata = age_data), 
           data = age_data, 
           palette = "uchicago",
           conf.int = T,      
           legend.labs = c("Age=25\nChemo=0", "Age=25\nChemo=1", 
                           "Age=75\nChemo=0", "Age=75\nChemo=1" ),
           break.x.by = 365,
           break.y.by = 0.1,
           surv.median.line = "hv")
```

Age matters a lot in probability of survival from breast cancer. Older
people have lower chance of living with or without chemo treatment
compared to younger people.

# Let's take tumor size into account as well and creating data for survplot

```{r}
size_data <- with(rott,                            
                  data.frame(age = rep(mean(age),6),
                             chemo = c(0, 0, 0, 1, 1, 1),
                             size =  c(1, 2, 3, 1, 2, 3)))

# Creating the survplot
ggsurvplot(survfit(cox_m1, 
                   newdata = size_data), 
           data = size_data, 
           conf.int = T,  
           palette = "uchicago",
           legend.labs = c("Chemo=0\nSize=1", "Chemo=0\nSize=2", "Chemo=0\nSize=3",
                         "Chemo=1\nSize=1", "Chemo=1\nSize=2", "Chemo=1\nSize=3"),
           break.x.by = 365,
           break.y.by = 0.1,
           surv.median.line = "hv")
```

Size of the tumor matter a lot as well! Bigger tumors are likely to lead
to lower survival probability.

# Checking the proportional hazard assumption

In principle, the Schoenfeld residuals are independent of time. A plot
that shows a non-random pattern against time is evidence of violation of
the PH assumption.

If a covariate breaks the assumption, it is time dependent. So, if we
cannot reject the null (p \< 0.05), our proportional hazards assumption
is reasonable, and our coefficientsdo not exhibit enough tie dependent
behavior to warrant concern. A very small p-values indicates a dependent
coefficients and violation of the assumption.

```{r}
cox.zph(cox_m1)
```

Systematic departures from a horizontal line are indicative of
non-proportional hazards, since proportional hazards assumes that
estimates β1,β2,β3 do not vary much over time.

```{r}
ggcoxzph(cox.zph(cox_m1))
```
# Goodness of Fit 

To test influential observations or outliers, we can plot the martingale
residuals against linear predictions

```{r}
ggcoxdiagnostics(cox_m1)
```

Positive values correspond to individuals that “died too soon” compared
to expected survival times. Negative values correspond to individual
that “lived too long”. Very large or small values are outliers, which
are poorly predicted by the model.

# Testing non-linearity

While the Cox PH model is semi-parametric and makes no assumptions about the form of the baseline hazard, it does assume a linear relationship between continuous predictors of the log hazard. 

```{r}
ggcoxfunctional(cox_m1)
```

This might help to properly choose the functional form of continuous
variable in the Cox PH model. For a given continuous covariate, patterns
in the plot may suggest that the variable is not properly fit.
Non-linearity is not an issue for categorical variables, so we only
examine plots of martingale residuals and partial residuals against a
continuous variable. From the plot above, it appears that age may not have a linear relationship woth the hazard function. 

# Shifting to fully parameterized

Fully parameterized models specifiy the form of the baseline hazard  

```{r}
w1 <- survreg(status ~ chemo + size + age, data = rott, dist = "weibull")
e1 <- survreg(status ~ chemo + size + age, data = rott, dist = "exponential")

stargazer(w1, e1, type = "text")

stargazer(w1, e1 , 
          type = "text", 
          column.labels = c(w1$dist, e1$dist),
          add.lines=list(c("Log(scale)", signif(log(w1$scale),3), signif(log(e1$scale),3)),
                         c("Scale", signif(w1$scale,3), signif(e1$scale,3)),
                         c("")
          )
)
```

## Check AIC for model selection

Lower values indicate a better fit model. 

```{r}
AIC(cox_m1, w1, e1)
```

# Postestimation and validation

"Weibull distribution might appear curvilinear in the log versus
time plot, but will be linear in a loglog plot ln[− ln S(t)]"

"Exponential distribution will appear linear in both plots, and
have a slope equal to 1 in the loglog plot"

## Log(survival time) against time:

```{r}
sfit1 <- survfit(cox_m1) 
sum1 <- summary(sfit1, times = rott$dtime)
f1 <- log(surv) ~ (time)
t1 <- as.data.frame(sum1[c("time", "surv")])

fit.lm1 <- lm(f1, t1)
plot(f1, t1)
abline(fit.lm1)
```
Derivation from linearity suggests that hazard rate may not be constant over time (as assumed by exponential model) 

## Log(-log(survival time) against log(time):

```{r}

sfit2 <- survfit(cox_m1) 
sum2 <- summary(sfit2, times = rott$dtime)
f2 <- log(-log(surv)) ~ log(time)
t2 <- as.data.frame(sum2[c("time", "surv")])

# Add a small offset to prevent extreme survival values from causing issues
t2$surv <- ifelse(t2$surv == 1, 0.999999, t2$surv)
t2$surv <- ifelse(t2$surv == 0, 0.000001, t2$surv)

fit.lm1 <- lm(f2, t2)
plot(f2, t2)
abline(-10, 1)
```

## Comparison graph (from eha package)

Close alignment suggest Weibull is a good fit. 

```{r}
status <- Surv(rott$dtime, rott$death) 
rott$size <- as.numeric(rott$size) 

# Cox regression 
cox <- coxreg(status ~ meno + grade + age, data = rott) 
# PH regression, standard is weibull
phw <- phreg(status ~ meno + grade + age, data = rott)  

check.dist(cox, phw) 
```

## Weibull version

The function rweibull(n, shape, scale) generates a sample of 10,000 observations from a Weibull distribution. When the shape parameter is set to 1, it will converge to an exponential distribution. 

```{r}
sample1 <- as.data.frame(rweibull(10000, 1, 1))
names(sample1) <- "Weibull"
head(sample1)

sample1 %>% 
  ggplot(aes(x = Weibull)) +
  geom_density(size = 1.2, color = "red") + 
  theme_minimal() + 
  labs(x = "Values", y = "Density") 

```

## Exponential version

```{r}
sample2 <- as.data.frame(rexp(10000, 1))
names(sample2) <- "Exponential"
head(sample2)

sample2 %>% 
ggplot(aes(x = Exponential)) +
  geom_density(size = 1.2, color = "blue") + 
  theme_minimal() + 
  labs(x = "Values", y = "Density") 
```
